---
title: "Zillow Housing Analysis Report (2018-2023)"
author: "Brian Chang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
knitr::opts_knit$set(root.dir = '~/Excel')
```
### I. Objective

> Zillow is a popular American realtor marketplace tool that many homebuyers use to begin their search. The purpose of this analysis is to investigate the following questions:

1. What states/regions in the U.S. have experienced the most mean growth in housing cost?
2. Are there certain times of the year when housing prices grow/decline?
3. What regions have the lowest list price compared to rate of percentage growth?

### II. Preparation
##### Load libraries and import dataset

Source: https://www.zillow.com/research/data/

* Category: List and Sale Prices
* Data Type: Median List Price (Smooth, SFR Only, Weekly)
* Geography: Metro & U.S.
```{r echo=T, results='hide', message=F, warning=F}
library(tidyverse)
df_zil <- read_csv("Zillow_metro_mlp_uc_sfr_sm_week.csv")
```
```{r collapse=T}
head(df_zil)
```

### III. Processing
##### Transform to long data format
```{r echo=T, results='hide', message=F, warning=F}
df_zil_long <- df_zil %>% 
                    pivot_longer(cols=starts_with("20"),
                                 names_to="Date", 
                                 values_to="MedianPrice")
```
```{r}
colnames(df_zil_long)
str(df_zil_long)
```

##### Convert string dates to Date class
```{r echo=T, results='hide', message=F, warning=F}
df_zil_clean <- df_zil_long %>%
                      mutate(DateParsed = as.Date(Date,"%Y-%m-%d")) %>%
                      select(-one_of("Date"))
```
```{r}
str(df_zil_clean)
```

##### Check the date range of the dataset
```{r collapse=T}
min(df_zil_clean$DateParsed)
max(df_zil_clean$DateParsed)
```
##### Remove null Date values
```{r collapse=T}
colSums(is.na(df_zil_clean))
df_zil_clean <- df_zil_clean[complete.cases(df_zil_clean[,6]),]
colSums(is.na(df_zil_clean))
```

### IV. Analysis
##### Group regions together by state
```{r echo=T, results='hide', message=F, warning=F}
df_state <- df_zil_clean %>%
              group_by(StateName, DateParsed) %>%
              summarise(AvgPriceState = mean(MedianPrice)) #%>%
              #filter(StateName == "NJ")
```

```{r collapse=T}
head(df_state)
```

```{r}
ggplot(df_state, aes(x = DateParsed, y = AvgPriceState, color = StateName)) +
    geom_line() +
    labs(title="Increase of Housing List Prices", y="Mean List Price") +
    scale_y_continuous(labels=scales::comma) +
    theme(axis.title.x=element_blank())
```

##### Calculate change in price from start to end date
```{r echo=T, results='hide', message=F, warning=F}
df_state_diff <- df_state %>%
                  group_by(StateName) %>%
                  mutate(Start = AvgPriceState[which.min(DateParsed)],
                         End = AvgPriceState[which.max(DateParsed)],
                         Diff = End - Start) %>%
                  summarise(Start=mean(Start), End=mean(End), Diff=mean(Diff)) %>%
                  arrange(desc(Diff))
```

```{r collapse=T}
  head(df_state_diff)
```

##### i) What states/regions in the U.S. have experienced the most mean growth in housing cost?
```{r}
top_n(df_state_diff, n=20, Diff) %>%
ggplot(., aes(x=reorder(StateName,-Diff), y=Diff)) +
    geom_bar(stat="identity") +
    labs(title="Top 20 States by Price Increase", y="Price Difference") +
    scale_y_continuous(labels=scales::comma) +
    theme(axis.title.x=element_blank())
```

##### ii) Are there certain times of the year when housing prices grow/decline?
```{r}
df_month <- df_zil_clean %>% 
  filter(!is.na(StateName)) %>% 
  group_by(Month=month(DateParsed)) %>% 
  summarise(MeanPrice = mean(MedianPrice))

df_month %>% ggplot(., aes(x=Month, y=MeanPrice)) +
  geom_line() +
  labs(title="U.S. Mean List Price Change by Month") +
  scale_y_continuous(name="Mean Price", labels=scales::comma) +
  scale_x_continuous(name="Month", breaks=seq(1,12))
```

##### iii) What regions have the lowest list price compared to rate of percentage growth?
```{r}
df_percent <- df_zil_clean %>% 
    group_by(RegionName) %>%
    mutate(Start = MedianPrice[which.min(DateParsed)],
           End = MedianPrice[which.max(DateParsed)],
           Diff = End - Start,
           PcntInc = Diff / Start) %>%
    filter(End < 200000) %>% # current price under $x
    summarise(Start=mean(Start), End=mean(End), Diff=mean(Diff), PcntInc=mean(PcntInc)) %>%
    arrange(desc(PcntInc))
head(df_percent, 100)
```
### V. Conclusions

* Certain states (e.g. Massachusetts, Hawaii, Idaho) experienced high mean growth in housing costs over the past five years (>\$300,000,000). Meanwhile others (e.g. Mississippi, Louisiana, Illinois) grew significantly less (<\$50,000). 

* Listing prices tend to peak from mid-Spring to early-Summer (Apr-July), while dropping harshly during the winter season.

* It is difficult to find housing for under \$100,000 in 2023, with housing prices soaring well beyond that. However, there are plenty of houses under \$200,000 that have experienced 50% or more increase in mean growth over the past 5 years.
